---
title: "Portfolio Optimization & Forecast"
author: "Paula Wang"
date: "4/26/2021"
output:
  html_notebook:
    toc: yes
    toc_float: yes
  html_document:
    df_print: paged
    toc: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Abstract


The project is intended to optimize a portfolio that contains growth stocks, and interpret the optimized portfolios through efficient frontier. It will also predict the future performance of the optimized portfolio and the market index using the Monte Carlo simulation. By comparing the forecasted returns, we found that the portfolio's return has a the possibility to outperform the market index in the next 30 days. However, the predicted results are somewhat off compare to the real price data. This might because of the special post-pandemic timing, further study may include event study and behavioral finance to improve the model. 




### Introduction


The world's economy has performed poorly in the past year due to the pandemic, and lots of industries were affected by it. The year 2020 started strong, then the whole stock market experienced large shock when the pandemic first started in February, and was gradually recovering to the previous level during the summer. The market index S&P 500 has turned down almost 65% in March and finished the year up with 14% from the last year's level. Looking deeper into the growth stocks performance last year, people had high expectations for them because more people and work were moved online. Growth stocks(mainly technology stocks) are also the ones that carried the bear market back to normal, and also outperformed both the market and value stocks. Knowing this, we will construct a portfolio with growth stocks and optimize it based on its past performance. We will also forecast the return of the optimized portfolio to see if it still can outperform the market in the next 30 days. There are some analysts shifted their favor from growth stock to value stock, because they believe people are getting back to normal after vaccine came out, and the growth stock won't be as popular as last year. 

The previous researches within this area implemented different methodologies and focused on different topics. There is one research uses the meta-heuristic technique to optimize the portfolio, and the particle swarm optimization(PSO) can build optimal risky portfolio really well. It also uses the sharpe ratio as measurement index, but the PSO framework is highly depend on its parameters. There is also another research that combines the Progressive Hedging Algorithm, Maximum Principle and fixed-point iteration scheme together to solve the discrete time optimal control problem. It formulate the portfolio problem as a stochastic programming problem through the scenario decomposition approach, which can be easily solved as a discrete time optimal control problem. The result of this research can fully explore the the properties of the portfolio problem under different time frames. There are also researches that focus on practical portfolio selection with preferences and so on. In the portfolio management field, there are lots of different topics and methods that the research can be focus on. In this project, we will primarily use the efficient frontier and Monte Carlo simulation to optimize and forecast a given portfolio. 




### Methods


Methods overview: The investment risk can be reduced and return can be maximized by applying portfolio optimization. The measurement indexes used in this process include: log returns, covariance matrix, and sharpe ratio. These measurements will give us an overall insight of the stock prices. To forecast the future performance of the portfolio, we will use Monte Carlo Simulation. It will take the stocks' past risk and return as reference, to predict the portfolio's return as a whole. 

Next, we will look at more details and go through the methods and codes that are used in this project. 

```{r}
library(plotly)
library(timetk)
library(tidyverse)
library(ggplot2)
library(broom)
library(dplyr)
```

```{r}
if (!require("quantmod")) {
    install.packages("quantmod")
    library(quantmod)
}
```

#### Portfolio Optimization

```{r}
start <- as.Date("2020-03-21")
end <- as.Date("2021-03-21")
```

```{r}
# Portfolio Components: Netflix, Paypal, Shopify, Google, Facebook, Alibaba
getSymbols("NFLX", src = "yahoo", from = start, to = end)
getSymbols("PYPL", src = "yahoo", from = start, to = end)
getSymbols("SHOP", src = "yahoo", from = start, to = end)
getSymbols("GOOGL", src = "yahoo", from = start, to = end)
getSymbols("FB", src = "yahoo", from = start, to = end)
getSymbols("BABA", src = "yahoo", from = start, to = end)
NFLX <- NFLX[, "NFLX.Close"]
PYPL <- PYPL[, "PYPL.Close"]
SHOP <- SHOP[, "SHOP.Close"]
GOOGL <- GOOGL[, "GOOGL.Close"]
FB <- FB[, "FB.Close"]
BABA <- BABA[, "BABA.Close"]
```

```{r}
#Combine all the closed stock prices together
stocks <- cbind(NFLX,PYPL,SHOP,GOOGL,FB,BABA)
head(stocks)
```

The codes above imported six stocks close price data (Netflix, Paypal, Shopify, Google, Facebook, Alibaba) from Yahoo.Finance and combined them all into one dataframe. The data is range from 3/21/2020 - 3/21/2021, and has a one year range in total. 

```{r}
#Compute the daily log return
stocks_change = stocks %>% log %>% diff
head(stocks_change)
```

```{r}
#Compute column means and covariance matrix
mean_change <- colMeans(stocks_change, na.rm=TRUE)
print(round(mean_change,5))
```

Then we compute the daily log returns. Log returns are frequently used in Finance. A lognormal distribution should be created when there is a need to calculate continuously compounded returns. The values within the distribution are always positive and future stock price generated from it will always be positive as well.(Because the stock price cannot fall below zero) The first row is NA because it doesn't have any row ahead of it, so the calculation won't generate any result. We also compute the average daily returns here which is the column means of the daily return for each stock.

```{r}
#Compute covariance matrix
cov_mat <- cov(stocks_change[-c(1),]) * 252
print(round(cov_mat,4))
```

Covariance matrix is calculated from the daily stock returns, it gets rid of the first row that contains no value and annualized it.(There are about 252 trading days each year) It can show the degree of correlations of different stocks' annual return. Typically, the correlations among stocks in the portfolio are expected to be low. The more diverse is the stocks, the less risky is the portfolio. When there are unexpected downturns in one stock, it won't affect other stocks in the portfolio by much.

```{r}
#Assign random numbers between 0 and 1 to each stocks
set.seed(1)
tick <- c('NFLX', 'PYPL', 'SHOP', 'GOOGL', 'FB', 'BABA')
wts <- runif(n = length(tick))
print(wts)
```

```{r}
#Shift the randomly assigned numbers into weights for different stocks
wts2 <- wts/sum(wts)
print(wts2)
```

```{r}
#Calculate the annualized portfolio return given the new weights
port_returns <- (sum(wts2 * mean_change) + 1)^252 - 1
port_returns
```

```{r}
#Calculate the portfolio risk by a series of matrix multiplications
port_risk <- sqrt(t(wts2) %*% (cov_mat %*% wts2))
print(port_risk)
```

```{r}
# Calculate the sharpe ratio using the return and risk calculated above
sharpe_ratio <- port_returns/port_risk
print(sharpe_ratio)
```

The code above asked the system to randomly generate numbers between 0 and 1 to each stocks. Next, we reform these random numbers into weights of the portfolio.(that can sum up to 1) Using the new weights created, we'll calculate the annualized portfolio return by multiplying the weights by average daily returns and minus one(the original investment) in the end. The portfolio risk is calculated by the squre root of transpose of weights * (covariance matrix*weights). The "covariance matrix times weights" gives the average daily stock returns under the different weights, and the square root of the transpose of the weights times the result of it reduce the dimension of the matrix. This will give us an understanding of how stocks return are correlated under the weights assigned and eventually give a number that represents the risk of the portfolio. The sharpe ratio is calculated as the return divided by risk, assume a zero risk free return here. It represents how much return is receiving for bearing one percent of the risk. Therefore, the higher sharpe ratio, the more effective the investment is.

```{r}
#Set the number of times of simulation
num_port <- 5000
```

```{r}
#Weights of all the randomly generated portfolio
all_wts <- matrix(nrow = num_port,
                  ncol = length(tick))
```

```{r}
#Returns of all the randomly generated portfolio
port_returns <- vector('numeric', length = num_port)
```

```{r}
#Risk of all the randomly generated portfolio
port_risk <- vector('numeric', length = num_port)
```

```{r}
#Sharpe ratio all the randomly generated portfolio
sharpe_ratio <- vector('numeric', length = num_port)
```

```{r}
#Repeat the last step 5000 times and generate 5000 different weights(portfolios)
for (i in seq_along(port_returns)) {
  
  wts <- runif(length(tick))
  wts <- wts/sum(wts)
  
  # Storing weight in the matrix
  all_wts[i,] <- wts
  
  # Portfolio returns
  port_ret <- sum(wts * mean_change)
  port_ret <- ((port_ret + 1)^252) - 1
  
  # Storing Portfolio Returns values
  port_returns[i] <- port_ret
  
  
  # Creating and storing portfolio risk
  port_sd <- sqrt(t(wts) %*% (cov_mat  %*% wts))
  port_risk[i] <- port_sd
  
  # Creating and storing Portfolio Sharpe Ratios
  # Assuming 0% Risk free rate
  sr <- port_ret/port_sd
  sharpe_ratio[i] <- sr
  
}
```

Here, we declared some variables to store the values generated within the process and repeated the last step 5000 times. 

```{r}
# Storing the values in the table
portfolio_values <- tibble(Return = port_returns,
                  Risk = port_risk,
                  SharpeRatio = sharpe_ratio)


# Converting matrix to a tibble and changing column names
all_wts <- tk_tbl(all_wts)
```

```{r}
colnames(all_wts) <- colnames(stocks_change)

# Combing all the values together
portfolio_values <- tk_tbl(cbind(all_wts, portfolio_values))
```

```{r}
head(portfolio_values)
```

Then we put all the outputs in the same dataframe so that it will be ready to analysis.

```{r}
min_var <- portfolio_values[which.min(portfolio_values$Risk),]
max_sr <- portfolio_values[which.max(portfolio_values$SharpeRatio),]
```

```{r}
p <- portfolio_values %>%
  ggplot(aes(x = Risk, y = Return, color = SharpeRatio)) +
  geom_point(size=0.5) +
  theme_classic() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = 'Annualized Risk',
       y = 'Annualized Returns',
       title = "Portfolio Optimization & Efficient Frontier") +
  geom_point(aes(x = Risk,
                 y = Return), data = min_var, color = 'red') +
  geom_point(aes(x = Risk,
                 y = Return), data = max_sr, color = 'red') +
  annotate('text', x = 0.34, y = 0.50, label = "Minimum variance portfolio") +
  annotate('text', x = 0.35, y = 1.54, label = "Tangency Portfolio") 

ggplotly(p)
```

Here we listed the portfolio with minimum variance(smallest risk) and maximum sharpe ratio(most effective). And we also have all the 5000 generated portfolio plotted on the efficient frontier with the annulized risk on the x-axis and annualized return on the y-axis.

#### Monte Carlo Simulations

```{r}
# Number of Monte Carlo Simulations and the training days
mc_rep = 1000
training_days = 30 
```

```{r}
# Define a function that calculate the ordinary return of the data
returns = function(Y){
  len = nrow(Y)
  yDif = unclass(Y[2:len, ]) / unclass(Y[1:len-1, ]) - 1
}

# Get the Stock Returns
stock_Returns = returns(stocks)
head(stock_Returns)
```

```{r}
# Get the Variance Covariance Matrix of Stock Returns
coVarMat = cov(stock_Returns)
miu = colMeans(stock_Returns)
# Extend the vector to a matrix
Miu = matrix(rep(miu, training_days), nrow = 6)
```

```{r}
#Take the generated minimum variance portfolio
vector1 <- c(0.2144948,0.01176868,0.03982831,0.391175,0.04915577,0.2935774)
portfolio_Weights <- array(c(vector1),dim = c(1,6))
portfolio_Weights
```

The first step in the Monte Carlo Simulation is declare the necessary variables - the number of simulation times and training(forecast) days. We also calculated the daily returns of each stock, and the covariance matrix and column means(average daily return) that are generated from it. We repeated the stocks' average daily return 6 times, so that the matrix will be 6*30 for later calculation. We will take the portfolio that has the minimum variance from the portfolio optimization results above as the forecast portfolio here. 

```{r}
# Initializing simulated 30 day portfolio returns
portfolio_Returns_30_m = matrix(0, training_days, mc_rep)

set.seed(200)
for (i in 1:mc_rep) {
  Z = matrix ( rnorm( dim(stock_Returns)[2] * training_days ), ncol = training_days )
  # Lower Triangular Matrix from our Choleski Factorization
  L = t( chol(coVarMat) )
  # Calculate stock returns for each day
  daily_Returns = Miu + L %*% Z  
  # Calculate portfolio returns for 30 days
  portfolio_Returns_30 = cumprod( portfolio_Weights %*% daily_Returns + 1 )
  # Add it to the monte-carlo matrix
  portfolio_Returns_30_m[,i] = portfolio_Returns_30;
}
```

This chunk of code is the Monte Carlo Simulation. We first set the seed so that the result will be constant. The simulation will repeat 1000 times as we declared above, then we will add some noise to the daily returns calculated above. This is because the stock price always have small changes that are unpredictable, so we can add some random number generated from the normal distribution into the stock returns we already have. The 6*30 matrix we created above are used to cooperate with the matrix size of the transpose of the covariance matrix. We add the noise times the covariance matrix to the actual return to mimic the stock price changes. In the last we assign the weights generated to the fixed return and collect the results. 

```{r}
# Visualising result
x_axis = rep(1:training_days, mc_rep)
y_axis = as.vector(portfolio_Returns_30_m-1)
plot_data = data.frame(x_axis, y_axis)
ggplot(data = plot_data, aes(x = x_axis, y = y_axis)) + geom_path(col = 'red', size = 0.1) +
  xlab('Days') + ylab('Portfolio Returns') + 
  ggtitle('Simulated Portfolio Returns in 30 days')+
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
Avg_Portfolio_Returns = mean(portfolio_Returns_30_m[30,]-1)
SD_Portfolio_Returns = sd(portfolio_Returns_30_m[30,]-1)
Median_Portfolio_Returns = median(portfolio_Returns_30_m[30,]-1)
print(c(Avg_Portfolio_Returns,SD_Portfolio_Returns,Median_Portfolio_Returns))
```

```{r}
Avg_CI = quantile(portfolio_Returns_30_m[30,]-1, c(0.025, 0.975))
print(Avg_CI)
```

In the end of the simulation, we visualize the forecasted 30 return in a plot. And we also get some insights from the data, such as average/standard deviation/median portfolio return and the confidence interval of the simulation. 

Next, we'll repeat the same process for SPY and forecast the future market performance.

```{r}
getSymbols("SPY", src = "yahoo", from = start, to = end)
SPY <- SPY[, "SPY.Close"]
```

```{r}
SPY_Return = returns(SPY)
head(SPY_Return)
```

```{r}
# Get the Variance Covariance Matrix of Stock Returns
coVarMat = cov(SPY_Return)
miu2 = colMeans(SPY_Return)
# Extend the vector to a matrix
Miu2 = matrix(rep(miu2, training_days), nrow = 1)
```

```{r}
# Initializing simulated 30 day portfolio returns
SPY_Returns_30_m = matrix(0, training_days, mc_rep)

set.seed(200)
for (i in 1:mc_rep) {
  Z = matrix ( rnorm( dim(SPY_Return)[2] * training_days ), ncol = training_days )
  # Lower Triangular Matrix from our Choleski Factorization
  L = t( chol(coVarMat2) )
  # Calculate stock returns for each day
  daily_Returns = Miu2 + L %*% Z  
  # Calculate portfolio returns for 30 days
  SPY_Returns_30 = cumprod( 1 %*% daily_Returns + 1 )
  # Add it to the monte-carlo matrix
  SPY_Returns_30_m[,i] = SPY_Returns_30;
}
```

```{r}
# Visualising result
x_axis = rep(1:training_days, mc_rep)
y_axis = as.vector(SPY_Returns_30_m-1)
plot_data = data.frame(x_axis, y_axis)
ggplot(data = plot_data, aes(x = x_axis, y = y_axis)) + geom_path(col = 'red', size = 0.1) +
  xlab('Days') + ylab('SPY Returns') + 
  ggtitle('Simulated SPY Returns in 30 days')+
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
Avg_SPY_Returns = mean(SPY_Returns_30_m[30,]-1)
SD_SPY_Returns = sd(SPY_Returns_30_m[30,]-1)
Median_SPY_Returns = median(SPY_Returns_30_m[30,]-1)
print(c(Avg_SPY_Returns,SD_SPY_Returns,Median_SPY_Returns))
```

```{r}
Avg_CI = quantile(SPY_Returns_30_m[30,]-1, c(0.025, 0.975))
print(Avg_CI)
```

#### Compare the result with real data

```{r}
#Calculate portfolio 30-day return
Portfolio_return <- 0.2144948*((508.9-523.11)/523.11) + 0.01176868*((264.89-244.38)/244.38) + 0.03982831*((1142.06-1160)/1160) + 0.391175*((2278.35-2030.69)/2030.69) + 0.04915577*((301.47-293.54)/293.54) + 0.2935774*((229.44-237.12)/237.12)
Portfolio_return
```

```{r}
#Calculate market(SPY) 30-day return
Market_return <- ((416.07-392.59)/392.59)
Market_return
```

Here we take the real data from Yahoo.Finance, and calculate the real return of the portfolio and the market in the next 30 days(3/21/2021 - 4/21/2021). 




### Results


Firstly, the average daily return of the stocks are all range from 0.1% to 0.45%. The covariance matrix shows that the highest correlation is 0.1889 between Paypal and Shopify, and the lowest correlation is 0.0587 between Alibaba and Netflix. The correlations of all the stocks are range from 0.0587 - 0.1889. For the first random weight we generated, the resulted annualized return is 1.1, risk is 0.35, and the sharpe ratio is 3.15.

Secondly, among all the simulated portfolios, the portfolio with the minimum variance is having a annualized return of 0.73, a risk of 0.3, and a sharpe ratio of 2.4. The portfolio with the maximum sharpe ratio is having a return of 1.6, a risk of 0.4, and a sharpe ratio of 3.8. All the simulated portfolios are put in the efficient frontier graph as a output. 

The confidence interval of the forecasted portfolio return in the next 30 days is between -13.2% - 30.4%, with an average return of 7.02%. And the confidence interval of the forecasted market performance is between -8.9% - 25.4%, with an average return of 6.9%. While the real data shows that the portfolio is generating 3.41% return while the market is generating 5.98% return.




### Discussion


#### Portfolio Optimization


In this section we will interpret the results and the outputs that are generated from the portfolio optimization process. In the beginning, the stocks' average daily returns are all positive and shows that these stocks are performed well during the last year. The output of the covariance matrix indicates that the overall correlations are not high. Typically, correlations below 0.3 are considered low, and we can say that the portfolio's risk is diversified pretty well in this case. From the first random weight we generated, the portfolio has been increase its value more than one time within the past year, the price volatility is 0.35, and there is 3.15 times of return for bearing 1 time of the risk. This investment result is pretty effective as the sharpe ratio is considered good when it is larger than 1. 

Next, we'll look at our 5000 simulations. To better understand the outputs of the simulation, we will bring the efficient frontier graph into our discussion. 

```{r}
p <- portfolio_values %>%
  ggplot(aes(x = Risk, y = Return, color = SharpeRatio)) +
  geom_point(size=0.5) +
  theme_classic() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::percent) +
  labs(x = 'Annualized Risk',
       y = 'Annualized Returns',
       title = "Portfolio Optimization & Efficient Frontier") +
  geom_point(aes(x = Risk,
                 y = Return), data = min_var, color = 'red') +
  geom_point(aes(x = Risk,
                 y = Return), data = max_sr, color = 'red') +
  annotate('text', x = 0.34, y = 0.50, label = "Minimum variance portfolio") +
  annotate('text', x = 0.35, y = 1.54, label = "Tangency Portfolio") 

ggplotly(p)
```

The efficient frontier graph looks like a horizontal U-shape. And now we can clearly see the position of each portfolio on the graph. The minimum variance portfolio is on the most left, because it is having the smallest risk and is the safest among all the generated portfolios. It is the best choice for the investors who have low risk tolerance and want to minimize the risk. The tangency portfolio is the one with the maximum sharpe ratio, and can generate the most return for bearing 1% of the risk. This is the best portfolio for the investors that can bear risks and want to maximize the efficiency of their investment. However, all the portfolios that are on the top of each risk level are the optimized portfolio, because they are generating the highest return for bearing the same risk. They are also the most efficient portfolios and has the highest sharpe ratio among the portfolios sharing the same risk. 


#### Monte Carlo Simulation


For the Monte Carlo Simulation, we will assume the investor is risk aversion and use the minimum variance portfolio generated above. From confidence interval and average return outputs, the market is having a narrower range and an similar average than the optimized portfolio. Even though the portfolio has a higher upper bound than the market, it is more volatile and therefore we will go for the less risky one and say that the market will be more efficient and is more likely to perform better in the next 30 days.

Compare to the return calculated from real stock data, the portfolio is generating 3.41% return while the market is generating 5.98% return. This is somewhat off compare to the average return we forecasted. This might result from the shift of the market sentiment. People are investing back in the industries other than technology since the society is gradually getting back to normal and the trend of investing in technology faded away. The assumption we made in the beginning has become true and the portfolio failed to outperform the market. For the future researches, people can include the behavioral finance and event analysis to make the prediction more accurate. 




### Reference


Venkiah, Va¨ına T.L. Portfolio Optimization through Genetic Algorithm and Particle Swarm Optimization. 


Barro, Diana, and Elio Canestrelli. Dynamic Portfolio Optimization: Time Decomposition Using the Maximum Principle with a Scenario Approach. 


Zucchi, Kristina. Lognormal and Normal Distribution. 13 Dec. 2020, www.investopedia.com/articles/investing/102014/lognormal-and-normal-distribution.asp. 


Diego, Israel. “Projects and Coursework /&gt; 2 Monte Carlo Simulation of Stock Portfolio in R, Matlab, and Python, 20 Dec. 2018, israeldi.github.io/bookdown/_book/monte-carlo-simulation-of-stock-portfolio-in-r-matlab-and-python.html. 


Jackson, Anna-Louise. “2020 Stock Market in Review: A Year That Defied Expectations.” Forbes, Forbes Magazine, 11 Mar. 2021, www.forbes.com/advisor/investing/stock-market-year-in-review-2020/. 













